---
title: "Yamanaka_RNASeq_Comparisons_Between_Analyses"
output: html_document
date: "2025-12-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


#install.packages("renv")
##renv::init()        # Initialize project environment
##renv::snapshot()    # Save package versions to renv.lock
##renv::restore()     # Recreate exact environment elsewhere

## Renew GitHub Token
# You must have a GitHub Personal Access Token (PAT) stored securely.
# If not already configured, run this once manually before using the script:
#gitcreds_set()

# # Retrieve token from the gitcreds store
# token <- gitcreds::gitcreds_get()$password


#if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Set Bioc version to latest (3.22)
#BiocManager::install(version = "3.22")

#bioc_pkgs <- c(
#  "KEGGREST", "Biobase", "AnnotationDbi", "S4Vectors", "Biostrings",
#  "Seqinfo", "BiocVersion", "org.Mm.eg.db", "XVector", "IRanges", "BiocGenerics"
#)

#BiocManager::install(bioc_pkgs, update = TRUE, ask = FALSE)
#renv::install(c("here", "fs", "DT", "dplyr", "stringr", "tibble", "qs"))
#renv::install("enrichplot")
#renv::install("tibble")
#renv::install("ggrepel")
#renv::install("ggvenn")
#renv::install("limma")
#renv::settings$bioconductor.version("3.22")
#renv::snapshot()   # records exact versions in renv.lock

library(usethis)
library(renv)
library(gitcreds)
library(dplyr)
library(stringr)
library(DESeq2)
library(here)
library(fs)
library(ggplot2)
library(pheatmap)
library(reshape2)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(clusterProfiler)
library(qs)
library(enrichplot)
library(tibble)
library(ggrepel)
library(ggvenn)
library(patchwork)
library(limma)

#Use external function script
source("scripts/plotPCA2.R")
source("scripts/runGSEA.R")

```

Documentation of used code.
In this script I compare our results with the OSKM and OKSM mouse models, published data from Pico et al. 2025.

First I need to load the files.

```{r load results tables from former analyses}

res_multi_df <- qread("results/differential_expression/res_multi_df.qs")
res_3d_df <- qread("results/differential_expression/res_3d_df.qs")
res_2wks_df <- qread("results/differential_expression/res_2wks_df.qs")

### .. read it:
M <- read.table("data/processed_data/counts_data/Pico2025_raw_counts.txt",
                header = TRUE,
                sep = "\t",
                quote = ""
)

dim(M)

# .. display the first 6 rows:
dt <- DT::datatable(M |> head())
dt

#only keep columns for WT and for OSKM and OKSM mice
M <- M %>% select("gene_id", "gene_name", "AS_KID_1", "AS_KID_2", "AS_KID_3", "AS_KID_4", "AS_KID_5", "AS_KID_6", "AS_KID_7", "AS_KID_8", "AS_KID_9", "AS_KID_10", "AS_KID_11", "AS_KID_12")

# Set gene_id (or gene_name) as rownames, then remove both annotation columns
rownames(M) <- M$gene_id
M <- M[, !names(M) %in% c("gene_id", "gene_name")]

#convert the column names to all-lowercase
colnames(M) <- tolower(colnames(M))

#discard genes with a total count < 10
min_count <- 10
M <- M %>% filter(rowSums(across(where(is.numeric))) >= min_count)

#map matrix entries with fractional digits to the next higher integer value
M <- M %>% mutate(across(where(is.numeric), ceiling))

```


## Set up the Input for DESeq2

In order to run the DESeq2 analysis, we need a meta table containing all information of different groups and treatments.


```{r ColData}
# Sample names must match the count matrix column names
sample_names_P <- c("as_kid_1", "as_kid_2", "as_kid_3", "as_kid_4", "as_kid_5", "as_kid_6", "as_kid_7", "as_kid_8", "as_kid_9", "as_kid_10", "as_kid_11", "as_kid_12")

# Create metadata
colData_P <- data.frame(
  row.names = sample_names,
  strain = c("WT", "WT", "WT", "4Fj", "4Fj", "4Fj", "4FsB", "4FsB", "4FsB", "4Fk", 
               "4Fk", "4Fk")
)


# Make sure group/treatment are factors
colData_P$strain <- factor(colData$strain)

#Set reference levels
colData_P$strain <- relevel(colData$strain, ref = "WT")

```

```{r DESeq2 analysis for Pico data}
## Running DESeq2 Analysis for each timepoint
# ============================================

# 3d Analysis
dds_Pico <- DESeqDataSetFromMatrix(
  countData = M,
  colData = colData_P,
  design = ~ strain  # or ~ genotype, if correction for age not wanted
)
dds_Pico <- DESeq(dds_Pico)

# compute all pairwise results separately
res_4Fj  <- as.data.frame(results(dds_Pico, contrast = c("strain", "4Fj", "WT")))
res_4FsB <- as.data.frame(results(dds_Pico, contrast = c("strain", "4FsB", "WT")))
res_4Fk  <- as.data.frame(results(dds_Pico, contrast = c("strain", "4Fk", "WT")))

# rename columns
colnames(res_4Fj)  <- paste0(colnames(res_4Fj), "_4Fj_vs_WT")
colnames(res_4FsB) <- paste0(colnames(res_4FsB), "_4FsB_vs_WT")
colnames(res_4Fk)  <- paste0(colnames(res_4Fk), "_4Fk_vs_WT")

# combine
res_all <- cbind(res_4Fj, res_4FsB, res_4Fk)
qsave(res_all, "results/differential_expression/res_all_Pico_Experiment.qs")

```


```{r PCA}

vsd_blind_Pico <- vst(dds_Pico, blind=TRUE)

p <- plotPCA2(vsd_blind_Pico,
         intgroup = "strain",
         ntop=500,
         PCs=c(x=1,y=2),
         colour = group) 
p
#ggsave("results/final_figures/PCA_3d.pdf", plot = p, width = 6, height = 4)

```


```{r read dds}

dds_3d <- qread("~/Documents/Arbeit_Forschung/Yamanaka/Pax8_RNASeq/Yamanaka_rnaseq/results/differential_expression/dds_3d.qs")
dds_2wks <- qread("~/Documents/Arbeit_Forschung/Yamanaka/Pax8_RNASeq/Yamanaka_rnaseq/results/differential_expression/dds_2wks.qs")

```

```{r Correlation Heatmap between experiments}
# unfortunately not helpful at all
library(limma)
library(pheatmap)

# Schritt 1: VST
vsd_3d   <- vst(dds_3d, blind = TRUE)
vsd_2wks <- vst(dds_2wks, blind = TRUE)
vsd_Pico <- vst(dds_Pico, blind = TRUE)

# Schritt 2: Matrizen extrahieren
mat_3d   <- assay(vsd_3d)
mat_2wks <- assay(vsd_2wks)
mat_Pico <- assay(vsd_Pico)

# Ensembl -> Entrez
entrez_ids <- mapIds(org.Mm.eg.db,
                     keys    = rownames(mat_Pico),
                     column  = "ENTREZID",
                     keytype = "ENSEMBL",
                     multiVals = "first")

# NAs entfernen
entrez_ids <- entrez_ids[!is.na(entrez_ids)]
mat_Pico   <- mat_Pico[names(entrez_ids), ]
rownames(mat_Pico) <- entrez_ids


# Schritt 3: Gemeinsame Gene
common_genes <- Reduce(intersect, list(rownames(mat_3d), rownames(mat_2wks), rownames(mat_Pico)))
mat_3d   <- mat_3d[common_genes, ]
mat_2wks <- mat_2wks[common_genes, ]
mat_Pico <- mat_Pico[common_genes, ]

# Schritt 4: Zusammenführen
mat_combined <- cbind(mat_3d, mat_2wks, mat_Pico)

# Schritt 5: Batch-Korrektur (3d und 2wks = gleiches Experiment)
batch <- c(rep("Pax8_OKS", ncol(mat_3d) + ncol(mat_2wks)), 
           rep("Pico", ncol(mat_Pico)))
mat_corrected <- removeBatchEffect(mat_combined, batch = batch)

# Schritt 6: Annotation
annotation_col <- data.frame(
  Experiment = batch,
  Group      = c(as.character(dds_3d$genotype), 
                 as.character(dds_2wks$genotype), 
                 as.character(dds_Pico$strain)),
  Timepoint  = c(rep("3d",   ncol(mat_3d)), 
                 rep("2wks", ncol(mat_2wks)), 
                 rep("Pico", ncol(mat_Pico)))
)
rownames(annotation_col) <- colnames(mat_corrected)

# Schritt 7: Correlation Heatmap
cor_matrix <- cor(mat_corrected)

pheatmap(cor_matrix,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Sample Correlation Heatmap")
```


Try a new heatmap, now using the logFC

```{r heatmap with logFC}

# Ensembl -> Entrez für die Pico Ergebnisse

convert_ensembl_to_entrez <- function(res_df) {
  entrez_ids <- mapIds(org.Mm.eg.db,
                       keys      = rownames(res_df),
                       column    = "ENTREZID",
                       keytype   = "ENSEMBL",
                       multiVals = "first")  # nimmt nur den ersten Treffer
  
  # NAs entfernen
  valid <- !is.na(entrez_ids)
  res_df     <- res_df[valid, ]
  entrez_ids <- entrez_ids[valid]
  
  # Duplikate entfernen (den mit niedrigstem padj behalten)
  dup <- duplicated(entrez_ids)
  res_df     <- res_df[!dup, ]
  entrez_ids <- entrez_ids[!dup]
  
  rownames(res_df) <- entrez_ids
  return(res_df)
}

res_4Fj  <- convert_ensembl_to_entrez(res_4Fj)
res_4FsB <- convert_ensembl_to_entrez(res_4FsB)
res_4Fk  <- convert_ensembl_to_entrez(res_4Fk)

# Nur log2FoldChange Spalte behalten
# common_genes neu berechnen basierend auf den results data frames
common_genes <- Reduce(intersect, list(rownames(res_3d), 
                                       rownames(res_2wks), 
                                       rownames(res_4Fj), 
                                       rownames(res_4FsB), 
                                       rownames(res_4Fk)))
length(common_genes)

common_genes <- Reduce(intersect, list(
  rownames(res_3d), 
  rownames(res_2wks), 
  rownames(res_4Fj), 
  rownames(res_4FsB), 
  rownames(res_4Fk)
))
length(common_genes)

lfc_combined <- data.frame(
  LFC_3d   = res_3d[common_genes,   "log2FoldChange"],
  LFC_2wks = res_2wks[common_genes, "log2FoldChange"],
  LFC_4Fj  = res_4Fj[common_genes,  "log2FoldChange_4Fj_vs_WT"],
  LFC_4FsB = res_4FsB[common_genes, "log2FoldChange_4FsB_vs_WT"],
  LFC_4Fk  = res_4Fk[common_genes,  "log2FoldChange_4Fk_vs_WT"],
  row.names = common_genes
)
dim(lfc_combined)

# Signifikante Gene aus mindestens einem Vergleich
sig_genes <- Reduce(union, list(
  rownames(res_3d[!is.na(res_3d$padj)     & res_3d$padj < 0.05, ]),
  rownames(res_2wks[!is.na(res_2wks$padj) & res_2wks$padj < 0.05, ]),
  rownames(res_4Fj[!is.na(res_4Fj$padj_4Fj_vs_WT)   & res_4Fj$padj_4Fj_vs_WT < 0.05, ]),
  rownames(res_4FsB[!is.na(res_4FsB$padj_4FsB_vs_WT) & res_4FsB$padj_4FsB_vs_WT < 0.05, ]),
  rownames(res_4Fk[!is.na(res_4Fk$padj_4Fk_vs_WT)   & res_4Fk$padj_4Fk_vs_WT < 0.05, ])
))

sig_genes <- intersect(sig_genes, common_genes)
length(sig_genes)

# Heatmap Matrix
lfc_heatmap <- lfc_combined[sig_genes, ]
dim(lfc_heatmap)

# Zusätzlicher LFC Filter
lfc_filter <- apply(lfc_heatmap, 1, function(x) any(abs(x) > 2.5, na.rm = TRUE))
lfc_heatmap_filtered <- lfc_heatmap[lfc_filter, ]
dim(lfc_heatmap_filtered)

# Gewünschte Spaltenreihenfolge definieren
col_order <- c("LFC_4FsB", "LFC_4Fj", "LFC_4Fk", "LFC_3d", "LFC_2wks")


pheatmap(lfc_heatmap_filtered,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-3, 3, length.out = 101),  # LFC Skala begrenzen
         show_rownames = FALSE,  # bei 1265 Genen nicht lesbar
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         main = "Log2FoldChange Heatmap (induced vs. control)")

```


Using now from Humphrey's single cell atlas the healthy mouse set.

```{r Humphrey's gene list}
gene_list <- c("Slc5a12", "Nox4", "Slc34a1", "Fut9", "Dab2", "Slc4a4", "Nhs", "Ndrg1", 
               "Slc13a1", "D630023F18Rik", "Lrp2", "Bnc2", "Slc6a19", "Gatm", "Ptprd", 
               "Cryl1", "Maf", "Auts2", "Abcc2", "Ano3", "Kap", "Cyp2e1", "Slc7a13", 
               "Cyp4b1", "Slc27a2", "Inmt", "Gpx1", "Ces1f", "Cndp2", "Cyb5a", "Selenop", 
               "Napsa", "Nudt19", "Ugt2b38", "Acy3", "Fth1", "Cd36", "Dbi", "Akr1c21", "Atp11a")

# Gensymbole -> Entrez IDs
entrez_list <- mapIds(org.Mm.eg.db,
                      keys      = gene_list,
                      column    = "ENTREZID",
                      keytype   = "SYMBOL",
                      multiVals = "first")

entrez_list <- entrez_list[!is.na(entrez_list)]

# Nur Gene die auch in lfc_combined vorhanden sind
entrez_list <- entrez_list[entrez_list %in% rownames(lfc_combined)]
length(entrez_list)

lfc_genelist <- lfc_combined[entrez_list, ]

# Zeilennamen zurück zu Gensymbolen für lesbare Heatmap
rownames(lfc_genelist) <- names(entrez_list)

# Gewünschte Spaltenreihenfolge definieren
col_order <- c("LFC_4FsB", "LFC_4Fj", "LFC_4Fk", "LFC_3d", "LFC_2wks")


p <- pheatmap(lfc_genelist[, col_order],
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-1.5, 1.5, length.out = 101),  # engere Skala
         show_rownames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         fontsize_row = 8,
         cellheight = 12,
         cellwidth = 40,
         main = "Log2FoldChange - Kidney reprogramming signatures")

ggsave("results/working_figures/heatmap_genelist.pdf", 
       plot = p, 
       width = 8, height = 10)

```

Trying another gene list, adding some more specific Tubular marker genes and taking out some more ubiquitous genes

```{r}
gene_list_PT <- c("Slc5a2", "Slc34a1", "Lrp2", "Slc13a3", "Slc22a6", "Slc7a13", 
                  "Slc27a2", "Gatm", "Slc6a19", "Napsa", "Cubn", "Hnf4a", "Slc5a12", 
                  "Nox4", "Dab2", "Cd36", "Gpx1", "Cyp2e1", "Cyp4b1", "Inmt", 
                  "Cndp2", "Acy3", "Ces1f", "Selenop", "Nudt19", "Ugt2b38")

# Gensymbole -> Entrez IDs
entrez_PT <- mapIds(org.Mm.eg.db,
                    keys      = gene_list_PT,
                    column    = "ENTREZID",
                    keytype   = "SYMBOL",
                    multiVals = "first")

entrez_PT <- entrez_PT[!is.na(entrez_PT)]
entrez_PT <- entrez_PT[entrez_PT %in% rownames(lfc_combined)]

cat("Gene gefunden:", length(entrez_PT), "von", length(gene_list_PT), "\n")

# LFC Matrix für PT Gene
lfc_PT <- lfc_combined[entrez_PT, ]
rownames(lfc_PT) <- names(entrez_PT)

# Heatmap
p <- pheatmap(lfc_PT[, col_order],
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-1, 1, length.out = 101),
         show_rownames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         fontsize_row = 9,
         cellheight = 14,
         cellwidth = 50,
         main = "Log2FoldChange - Proximal Tubule Marker Genes")
ggsave("results/working_figures/heatmap_genelist_PT.pdf", 
       plot = p, 
       width = 8, height = 10)
```

Now the same for collecting duct cells - using a gene list, created by Claude

```{r collecting duct}
gene_list_CD <- c(
  # Principal Cells (PC)
  "Aqp2",      # Aquaporin-2 - klassischer PC Marker
  "Aqp3",      # Aquaporin-3
  "Aqp4",      # Aquaporin-4
  "Scnn1a",    # ENaC alpha
  "Scnn1b",    # ENaC beta
  "Scnn1g",    # ENaC gamma
  "Avpr2",     # Vasopressin Rezeptor
  "Fxyd4",     # FXYD4 - PC spezifisch
  "Notch2",    # Notch2 - PC Identität
  
  # Type A Intercalated Cells (A-IC)
  "Atp6v1b1",  # H+-ATPase - klassischer IC Marker
  "Atp6v0d2",  # H+-ATPase Untereinheit
  "Slc4a1",    # AE1 Anionenaustauscher - A-IC spezifisch
  "Kit",       # c-Kit - A-IC Oberflächenmarker
  "Adgrf5",    # A-IC spezifisch
  "Hepacam2",  # A-IC Oberflächenmarker
  
  # Type B Intercalated Cells (B-IC)
  "Slc26a4",   # Pendrin - B-IC spezifisch
  "Slc4a9",    # AE4 - B-IC
  
  # Pan-Collecting Duct
  "Calb1",     # Calbindin
  "Kcnj1",     # ROMK Kaliumkanal
  "Jag1"       # Notch Ligand - IC zu PC Transition
)

# Gensymbole -> Entrez IDs
entrez_CD <- mapIds(org.Mm.eg.db,
                    keys      = gene_list_CD,
                    column    = "ENTREZID",
                    keytype   = "SYMBOL",
                    multiVals = "first")

entrez_CD <- entrez_CD[!is.na(entrez_CD)]
entrez_CD <- entrez_CD[entrez_CD %in% rownames(lfc_combined)]

cat("Gene gefunden:", length(entrez_CD), "von", length(gene_list_CD), "\n")

# LFC Matrix
lfc_CD <- lfc_combined[entrez_CD, ]
rownames(lfc_CD) <- names(entrez_CD)

# Heatmap
p <- pheatmap(lfc_CD[, col_order],
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-1, 1, length.out = 101),
         show_rownames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         fontsize_row = 9,
         cellheight = 14,
         cellwidth = 50,
         main = "Log2FoldChange - Collecting Duct Marker Genes")
ggsave("results/working_figures/heatmap_genelist_CD.pdf", 
       plot = p, 
       width = 8, height = 10)
```

and for glomerular cells

```{r Glomerular cells}
gene_list_Podo <- c(
  # Klassische Podocyten-Marker
  "Nphs1",     # Nephrin - der wichtigste Podocyten-Marker
  "Nphs2",     # Podocin
  "Wt1",       # WT1 Transkriptionsfaktor
  "Podxl",     # Podocalyxin
  "Synpo",     # Synaptopodin
  "Cd2ap",     # CD2-assoziiertes Protein
  
  # Glomeruläre Endothelzellen
  "Pecam1",    # CD31
  "Ehd3",      # spezifisch für glomeruläre Endothelzellen
  
  # Mesangialzellen
  "Pdgfrb",    # PDGFR-beta
  "Des",       # Desmin
  
  # Parietal Epithelial Cells (PEC)
  "Cldn1",     # Claudin-1
  "Ncam1"      # NCAM1
)

# Gensymbole -> Entrez IDs
entrez_Podo <- mapIds(org.Mm.eg.db,
                      keys      = gene_list_Podo,
                      column    = "ENTREZID",
                      keytype   = "SYMBOL",
                      multiVals = "first")

entrez_Podo <- entrez_Podo[!is.na(entrez_Podo)]
entrez_Podo <- entrez_Podo[entrez_Podo %in% rownames(lfc_combined)]

cat("Gene gefunden:", length(entrez_Podo), "von", length(gene_list_Podo), "\n")

# LFC Matrix
lfc_Podo <- lfc_combined[entrez_Podo, ]
rownames(lfc_Podo) <- names(entrez_Podo)

# Heatmap
p <- pheatmap(lfc_Podo[, col_order],
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-1, 1, length.out = 101),
         show_rownames = TRUE,
         cluster_cols = FALSE,
         cluster_rows = TRUE,
         fontsize_row = 9,
         cellheight = 14,
         cellwidth = 50,
         main = "Log2FoldChange - Glomerular Marker Genes (Negative Control)")
         
ggsave("results/working_figures/heatmap_genelist_gloms.pdf", 
       plot = p, 
       width = 8, height = 10)         
```

