---
title: "Yamanaka_PT_Expression_Analysis"
output: html_document
date: "2025-12-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


#install.packages("renv")
##renv::init()        # Initialize project environment
##renv::snapshot()    # Save package versions to renv.lock
##renv::restore()     # Recreate exact environment elsewhere

## Renew GitHub Token
# You must have a GitHub Personal Access Token (PAT) stored securely.
# If not already configured, run this once manually before using the script:
#gitcreds_set()

# # Retrieve token from the gitcreds store
# token <- gitcreds::gitcreds_get()$password


#if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Set Bioc version to latest (3.22)
#BiocManager::install(version = "3.22")

#bioc_pkgs <- c(
#  "KEGGREST", "Biobase", "AnnotationDbi", "S4Vectors", "Biostrings",
#  "Seqinfo", "BiocVersion", "org.Mm.eg.db", "XVector", "IRanges", "BiocGenerics"
#)

#BiocManager::install(bioc_pkgs, update = TRUE, ask = FALSE)
#renv::install(c("here", "fs", "DT", "dplyr", "stringr", "tibble", "qs"))
#renv::install("enrichplot")
#renv::install("tibble")
#renv::install("ggrepel")
#renv::install("ggvenn")
#renv::install("patchwork")
#renv::install("readxl")
#renv::install("ggVennDiagram")
#renv::install("kableExtra")

#renv::settings$bioconductor.version("3.22")
#renv::snapshot()   # records exact versions in renv.lock



library(usethis)
library(renv)
library(gitcreds)
library(dplyr)
library(stringr)
library(DESeq2)
library(here)
library(fs)
library(ggplot2)
library(pheatmap)
library(reshape2)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(clusterProfiler)
library(qs)
library(enrichplot)
library(tibble)
library(ggrepel)
library(ggvenn)
library(patchwork)
library(readxl)
library(ggVennDiagram)
library(kableExtra)

#Use external function script
source("scripts/plotPCA2.R")
source("scripts/runGSEA.R")

```


Documentation of used code.
In this script I use a subset of genes that are 1. specific for proximal tubules (Chen 2021), or 2. expressed in tubules in scRNA-Seq data (Szusztak) and perform analysis on this subset of genes.

First I need to load my results files.

```{r load results tables from former analyses}

res_multi_df <- qread("results/differential_expression/res_multi_df.qs")
res_3d_df <- qread("results/differential_expression/res_3d_df.qs")
res_2wks_df <- qread("results/differential_expression/res_2wks_df.qs")

```

Now I need to load the proximal tubular datasets.

```{r load PT Expression datasets}
PT_spec_genes <- read.csv("results/cell-specific analyses/Top200_PT_specific_genes_summary.csv")
PT_spec_genes <- PT_spec_genes$Gene_symbol# make a vector from tibble

PT_all_genes <- read_excel("data/reference_data/Park2018_OnlyPTGenes_ALL.xlsx")
PT_all_genes_filtered <- PT_all_genes %>% filter(`PT cell percentage` >= 10)
PT_all_genes_filtered <- PT_all_genes_filtered$Gene_name

```

I will subset the 2 analyses of the different time points for the PT-specific genes.

```{r subset PT specific genes, as well as all PT expressed genes}
res_3d_PT <- res_3d_df [res_3d_df$gene_symbol %in% PT_spec_genes, ] #only PT data set, from 116 of Chen 2021 paper, 110 are expressed in our data set

res_2wks_PT <- res_2wks_df [res_2wks_df$gene_symbol %in% PT_spec_genes, ]

res_3d_PT_all <- res_3d_df [res_3d_df$gene_symbol %in% PT_all_genes_filtered, ]
res_2wks_PT_all <- res_2wks_df [res_2wks_df$gene_symbol %in% PT_all_genes_filtered, ]
```

# 1. PT specific genes
## PCA analysis PT specific genes

I want to start with PCA analysis.

```{r Preparing PCA}
#set "Universe" - defined as all genes that are actually expressed for 3d
uni_genes_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj)) |>
  dplyr::pull(gene_symbol) |>
  unique()

#upregulated genes 3d
de_genes_up_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes 3d
de_genes_dn_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

## looking at STRONG effects for 3d, here now only with log2FoldChange of 0.5
#upregulated genes
de_genes_up_strong_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes
de_genes_dn_strong_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < -0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()


### Now the same for our 2 weeks analysis
#set "Universe" - defined as all genes that are actually expressed for 2 weeks
uni_genes_2wks_PT_spec <- res_2wks_PT |>
  dplyr::filter(!is.na(padj)) |>
  dplyr::pull(gene_symbol) |>
  unique()

#upregulated genes 2 wks
de_genes_up_2wks_PT_spec <- res_2wks_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes 2 wks
de_genes_dn_2wks_PT_spec <- res_2wks_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < 0) |>
  dplyr::pull(gene_symbol) |>
  unique()


## looking at STRONG effects, log2FoldChange only set at 0.5!!
#upregulated genes
de_genes_up_strong_2wks_PT_spec <- res_2wks_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes
de_genes_dn_strong_2wks <- res_2wks_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < -0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()


```


Now I can run the PCA analysis.

```{r run PCA analysis for PT specific genes 1}

#for 3d timepoint

#ego_up_3d_PT_spec <- enrichGO(gene = de_genes_up_3d_PT_spec,
 #                         universe = uni_genes_3d_PT_spec,
  #                        OrgDb = org.Mm.eg.db,
   #                       keyType = "SYMBOL",
    #                      ont = "BP",
     #                     pAdjustMethod = "BH",
      #                    pvalueCutoff = 0.05,
       #                   qvalueCutoff = 0.2,
        #                  readable = TRUE)

#DT::datatable(head(ego_up_3d_PT_spec)) # yields no results

# ego_dn_3d_PT_spec <- enrichGO(gene = de_genes_dn_3d_PT_spec,
#                           universe = uni_genes_3d_PT_spec,
#                           OrgDb = org.Mm.eg.db,
#                           keyType = "SYMBOL",
#                           ont = "BP",
#                           pAdjustMethod = "BH",
#                           pvalueCutoff = 0.05,
#                           qvalueCutoff = 0.2,
#                           readable = TRUE)
# 
# DT::datatable(head(ego_dn_3d_PT_spec)) # yields no results
# 
# #------------------------------

#for 2wks timepoint


# ego_up_2wks_PT_spec <- enrichGO(gene = de_genes_up_2wks_PT_spec,
#                           universe = uni_genes_2wks_PT_spec,
#                           OrgDb = org.Mm.eg.db,
#                           keyType = "SYMBOL",
#                           ont = "BP",
#                           pAdjustMethod = "BH",
#                           pvalueCutoff = 0.05,
#                           qvalueCutoff = 0.2,
#                           readable = TRUE)
# 
# DT::datatable(head(ego_up_2wks_PT_spec)) # yields no results
# 
# ego_dn_2wks_PT_spec <- enrichGO(gene = de_genes_dn_2wks_PT_spec,
#                           universe = uni_genes_2wks_PT_spec,
#                           OrgDb = org.Mm.eg.db,
#                           keyType = "SYMBOL",
#                           ont = "BP",
#                           pAdjustMethod = "BH",
#                           pvalueCutoff = 0.05,
#                           qvalueCutoff = 0.2,
#                           readable = TRUE)
# 
# DT::datatable(head(ego_dn_2wks_PT_spec)) # yields no results
```

The PCA analysis yielded no results, assumingly due to low gene numbers. 
I will also try once without the divisions into up- or downregulated genes.

```{r PCA regardless of direction}
de_genes_3d_PT_spec <- res_3d_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05) |>
  dplyr::pull(gene_symbol) |>
  unique()

de_genes_2wks_PT_spec <- res_2wks_PT |>
  dplyr::filter(!is.na(padj), padj <= 0.05) |>
  dplyr::pull(gene_symbol) |>
  unique()

# ego_3d_PT_spec <- enrichGO(gene = de_genes_3d_PT_spec,
#                           universe = uni_genes_3d_PT_spec,
#                           OrgDb = org.Mm.eg.db,
#                           keyType = "SYMBOL",
#                           ont = "BP",
#                           pAdjustMethod = "BH",
#                           pvalueCutoff = 0.05,
#                           qvalueCutoff = 0.2,
#                           readable = TRUE)
# 
# DT::datatable(head(ego_3d_PT_spec)) # yields no results
# 
# ego_2wks_PT_spec <- enrichGO(gene = de_genes_2wks_PT_spec,
#                           universe = uni_genes_2wks_PT_spec,
#                           OrgDb = org.Mm.eg.db,
#                           keyType = "SYMBOL",
#                           ont = "BP",
#                           pAdjustMethod = "BH",
#                           pvalueCutoff = 0.05,
#                           qvalueCutoff = 0.2,
#                           readable = TRUE)
# 
# DT::datatable(head(ego_2wks_PT_spec)) # yields no results

```

Also this analysis yielded no results.

## Venn diagrams for PT specific gene expression

Now I will look at Venn diagrams.

```{r Venn diagram for upregulated genes}
## Venn-Diagram Upregulated Genes
gene_list_up <- list(
  "Short term effect" = de_genes_up_3d_PT_spec,
  "Long term effect" = de_genes_up_2wks_PT_spec
)

short <- gene_list_up[["Short term effect"]]
long <- gene_list_up[["Long term effect"]]

# Calculate regions
only_short <- setdiff(short, long)
only_long <- setdiff(long, short)
both <- intersect(short, long)

# Create summary table
venn_genes <- tibble(
  Region = c("Only Short term effect", "Only Long term effect", "Both effects"),
  Count = c(length(only_short), length(only_long), length(both)),
  Genes = list(only_short, only_long, both)
 )

# preparing figure, using subtitle
venn <- ggvenn(gene_list_up, 
               fill_color = c("skyblue", "pink"),
               stroke_size = 0.5,
               set_name_size = 6.5,
               text_size = 6,
               show_percentage = FALSE) +
  ggtitle("Upregulated PT-specific Genes Overlap",
          subtitle = "CrePos vs Ctrl (padj < 0.05)") +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 14))


venn
ggsave("results/final_figures/venn_upregulated_PT_genes.pdf", plot = venn, width = 8, height = 6, units = "in", dpi = 300)



  # After calculating venn_genes, use this:
dt <- DT::datatable(
  venn_genes |> 
    as.data.frame(),  # Keep long strings
  options = list(
    scrollY = "300px", 
    scrollX = "100%", 
    pageLength = 10,
    autoWidth = FALSE,
    columnDefs = list(
      list(width = "60%", targets = 2),  # Wide genes column
      list(className = "dt-center", targets = 1)  # Center count
    )
  ),
  escape = FALSE,
  rownames = FALSE
)
dt

## Venn-Diagram Downregulated Genes
gene_list_dn <- list(
  "Short term effect" = de_genes_dn_3d_PT_spec,
  "Long term effect" = de_genes_dn_2wks_PT_spec
)

short <- gene_list_dn[["Short term effect"]]
long <- gene_list_dn[["Long term effect"]]

# Calculate regions
only_short <- setdiff(short, long)
only_long <- setdiff(long, short)
both <- intersect(short, long)

# Create summary table
venn_genes <- tibble(
  Region = c("Only Short term effect", "Only Long term effect", "Both effects"),
  Count = c(length(only_short), length(only_long), length(both)),
  Genes = list(only_short, only_long, both)
 )

# using subtitle
venn <- ggvenn(gene_list_dn, 
               fill_color = c("skyblue", "pink"),
               stroke_size =0.5,
               set_name_size = 6.5,
               text_size = 6,
               show_percentage = FALSE) +
  ggtitle("Downregulated PT-specific Genes Overlap",
          subtitle = "CrePos vs Ctrl (padj < 0.05)") +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 14))

venn
ggsave("results/final_figures/venn_downregulated_PT_genes.pdf", plot = venn, width = 8, height = 6, units = "in", dpi = 300)

# After calculating venn_genes, use this:
dt <- DT::datatable(
  venn_genes |> 
    as.data.frame(),  # Keep long strings
  options = list(
    scrollY = "300px", 
    scrollX = "100%", 
    pageLength = 10,
    autoWidth = FALSE,
    columnDefs = list(
      list(width = "60%", targets = 2),  # Wide genes column
      list(className = "dt-center", targets = 1)  # Center count
    )
  ),
  escape = FALSE,
  rownames = FALSE
)
dt

```

# 2. all PT - expressed genes (Park, Szusztak 2018)

```{r Preparing PCA for all PT expressed genes}
#set "Universe" - defined as all genes that are actually expressed for 3d
uni_genes_3d_PT_all <- res_3d_PT_all |>
  dplyr::filter(!is.na(padj)) |>
  dplyr::pull(gene_symbol) |>
  unique()

#upregulated genes 3d
de_genes_up_3d_PT_all <- res_3d_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes 3d
de_genes_dn_3d_PT_all <- res_3d_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

## looking at STRONG effects for 3d, here now only with log2FoldChange of 0.5
#upregulated genes
de_genes_up_strong_3d_PT_all <- res_3d_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes
de_genes_dn_strong_3d_PT_all <- res_3d_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < -0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()


### Now the same for our 2 weeks analysis
#set "Universe" - defined as all genes that are actually expressed for 2 weeks
uni_genes_2wks_PT_all <- res_2wks_PT_all |>
  dplyr::filter(!is.na(padj)) |>
  dplyr::pull(gene_symbol) |>
  unique()

#upregulated genes 2 wks
de_genes_up_2wks_PT_all <- res_2wks_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes 2 wks
de_genes_dn_2wks_PT_all <- res_2wks_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < 0) |>
  dplyr::pull(gene_symbol) |>
  unique()


## looking at STRONG effects, log2FoldChange only set at 0.5!!
#upregulated genes
de_genes_up_strong_2wks_PT_all <- res_2wks_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange > 0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()

#downregulated genes
de_genes_dn_strong_2wks_all <- res_2wks_PT_all |>
  dplyr::filter(!is.na(padj), padj <= 0.05, log2FoldChange < -0.5) |>
  dplyr::pull(gene_symbol) |>
  unique()


```

Now I can run the PCA analysis.

```{r run PCA analysis for PT specific genes}

#for 3d timepoint

ego_up_3d_PT_all <- enrichGO(gene = de_genes_up_3d_PT_all,
                          universe = uni_genes_3d_PT_all,
                          OrgDb = org.Mm.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.2,
                          readable = TRUE)

DT::datatable(head(ego_up_3d_PT_all)) # yields no results

ego_dn_3d_PT_all <- enrichGO(gene = de_genes_dn_3d_PT_all,
                          universe = uni_genes_3d_PT_all,
                          OrgDb = org.Mm.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.2,
                          readable = TRUE)

DT::datatable(head(ego_dn_3d_PT_all)) 

#Visualize

dotplot(ego_dn_3d_PT_all, showCategory=10) +
  ggtitle("GOEA - Downregulated Genes after 3 days")


#------------------------------

#for 2wks timepoint


ego_up_2wks_PT_all <- enrichGO(gene = de_genes_up_2wks_PT_all,
                          universe = uni_genes_2wks_PT_all,
                          OrgDb = org.Mm.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.2,
                          readable = TRUE)

DT::datatable(head(ego_up_2wks_PT_all)) 

ego_dn_2wks_PT_all <- enrichGO(gene = de_genes_dn_2wks_PT_all,
                          universe = uni_genes_2wks_PT_all,
                          OrgDb = org.Mm.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.2,
                          readable = TRUE)

DT::datatable(head(ego_dn_2wks_PT_all)) 

#Visualize
dotplot(ego_up_2wks_PT_all, showCategory=10) +
  ggtitle("GOEA - Upregulated Genes after 2 weeks")
dotplot(ego_dn_2wks_PT_all, showCategory=10) +
  ggtitle("GOEA- Downregulated Genes after 2 weeks")

```


```{r PT-Score}
## First, add gene names to the gene counts table
#read cleaned gene counts table
genecount_file_cleaned <- readRDS("data/processed_data/gene_counts_YFproject_cleaned.rds")

# Convert Entrez IDs of my res_results to gene symbols
entrez_ids <- rownames(genecount_file_cleaned)
gene_symbols <- mapIds(org.Mm.eg.db,
                       keys = entrez_ids,
                       column = "SYMBOL",
                       keytype = "ENTREZID",
                       multiVals = "first")  # take first if multiple matches

#Add gene symbols to the results table
genecount_file_cleaned <- as.data.frame(genecount_file_cleaned) |>
  mutate(
    entrez_id   = rownames(genecount_file_cleaned),
    gene_symbol = gene_symbols[entrez_id]
  )

# Convert Entrez IDs of my res_results to gene names
gene_names <- mapIds(
  org.Mm.eg.db,
  keys      = entrez_ids,
  column    = "GENENAME",
  keytype   = "ENTREZID",
  multiVals = "first"
)

#Add gene names to the results table
genecount_file_cleaned <- genecount_file_cleaned|>
  mutate(gene_name = gene_names[entrez_id])


## 
# Normalisierung auf Kontrollen
control_samples <- c("Y340", "Y343", "Y344", "Y347", "Y350", "Y351", "Y352", "Y356")

pt_genes_available <- res_3d_PT$gene_symbol
# pt_genes_available2 <- res_2wks_PT$gene_symbol #not necessary, yields same result

# Z-score basierend nur auf Kontrollen
# 1. Zuerst die Daten für PT-Gene filtern
pt_data <- genecount_file_cleaned[genecount_file_cleaned$gene_symbol %in% pt_genes_available, ]

#Rownames as gene names
rownames(pt_data) <- pt_data$gene_symbol

# 2. Dann nur die Kontroll-Samples auswählen und Mittelwert berechnen
control_mean <- rowMeans(pt_data[, control_samples])

# 3. Standardabweichung für Kontrollen
control_sd <- apply(pt_data[, control_samples], 1, sd)

# Dann alle Samples relativ zu Kontrolle normalisieren
pt_score_all <- colMeans(
  (bulk_expr[pt_genes_available, ] - control_mean) / control_sd
)

# Dann alle Samples (inkl. OSKM) normalisieren
all_samples <- colnames(pt_data)[!colnames(pt_data) %in% c("gene_symbol", "entrez_id", "gene_name")] # alle außer Gen-Info
pt_score_zscore <- colMeans(
  (pt_data[, all_samples] - control_mean) / control_sd
)

# 1. Check Gene-Overlap

pt_genes_available <- res_3d_PT$gene_symbol
# pt_genes_available2 <- res_2wks_PT$gene_symbol #not necessary, yields same result

# 2. PT-Score berechnen
pt_score <- colMeans(bulk_data[pt_genes_available, ])

# 3. Normalisierung
pt_score_normalized <- scale(pt_score)

# Basis-Ansatz
pt_genes <- c("Acsm2", "Spp2", "Slc34a1", ...) # deine 190 detektierten Gene
pt_signature <- colMeans(bulk_expr[pt_genes, ])

# Robuster: Z-score normalisiert
pt_signature_zscore <- apply(bulk_expr[pt_genes, ], 2, function(x) {
  mean(scale(x))
})

# Teste ob dein Score mit einzelnen Top-Markern korreliert
# Korrelation nur für Kontroll-Samples
cor(pt_score_zscore[control_samples], 
    as.numeric(pt_data[pt_data$gene_symbol == "Lrp2", control_samples]))

top_markers <- c("Lrp2", "Slc5a2", "Slc34a1", "Pck1", "Fbp1")

correlations_ctrl <- sapply(top_markers, function(gene) {
  gene_expr <- as.numeric(pt_data[pt_data$gene_symbol == gene, control_samples])
  cor(pt_score_zscore[control_samples], gene_expr)
})

print(correlations_ctrl)

# Check: Varianz in den Kontroll-Samples
pt_score_ctrl <- pt_score_zscore[control_samples]
print(pt_score_ctrl)
sd(pt_score_ctrl)  # Wenn sehr klein (<0.1), dann wenig Variation

# Variation einzelner Gene
lrp2_ctrl <- as.numeric(pt_data[pt_data$gene_symbol == "Lrp2", control_samples])
print(lrp2_ctrl)
sd(lrp2_ctrl) / mean(lrp2_ctrl)  # Coefficient of variation

# Korrelation mit ALLEN Samples (inkl. OSKM)
top_markers <- c("Lrp2", "Slc5a2", "Slc34a1", "Pck1", "Fbp1")

correlations_all <- sapply(top_markers, function(gene) {
  gene_expr <- as.numeric(pt_data[pt_data$gene_symbol == gene, all_samples])
  cor(pt_score_zscore, gene_expr)
})

print(correlations_all)

# Visualisierung für Lrp2
# ALLES ZUSAMMEN ausführen (markieren und Ctrl+Enter)
lrp2_all <- as.numeric(pt_data[pt_data$gene_symbol == "Lrp2", all_samples])

plot(lrp2_all, pt_score_zscore,
     col = ifelse(names(pt_score_zscore) %in% control_samples, "blue", "red"),
     pch = 19, cex = 1.5,
     xlab = "Lrp2 Expression (counts)",
     ylab = "PT Signature Score (Z-score)",
     main = paste("Correlation (all):", round(cor(pt_score_zscore, lrp2_all), 2)))
legend("topright", c("Control", "OSKM"), col = c("blue", "red"), pch = 19)
abline(lm(pt_score_zscore ~ lrp2_all), col = "black", lwd = 2)


# Vergleiche Mittelwerte
# Kontrollen (blau):
mean(lrp2_all[names(pt_score_zscore) %in% control_samples])
mean(pt_score_zscore[control_samples])

# OSK (rot):
osk_samples <- names(pt_score_zscore)[!names(pt_score_zscore) %in% control_samples]
mean(lrp2_all[!names(pt_score_zscore) %in% control_samples])
mean(pt_score_zscore[osk_samples])

# OSKM 3 Tage samples
osk_3d_samples <-c("Y341", "Y342", "Y345", "Y346")
mean(lrp2_all[names(pt_score_zscore) %in% osk_3d_samples])
mean(pt_score_zscore[osk_3d_samples])

# OSKM 2 weeks samples
osk_2wks_samples <-c("Y348", "Y349", "Y353", "Y354", "Y355")
mean(lrp2_all[names(pt_score_zscore) %in% osk_2wks_samples])
mean(pt_score_zscore[osk_2wks_samples])


# Negativkontrolle: Sollte NICHT korrelieren
cor(pt_signature, bulk_expr["Aqp2", ])  # Collecting Duct Marker
cor(pt_signature, bulk_expr["Slc12a3", ])  # DCT Marker

# Option B: Funktionale Gruppen
transporters <- pt_genes[grepl("^Slc", pt_genes)]
metabolic <- pt_genes[pt_genes %in% c("Pck1", "Fbp1", "G6pc", ...)]

score_transport <- colMeans(bulk_expr[transporters, ])
score_metabolic <- colMeans(bulk_expr[metabolic, ])

# Welche PT-Funktionen gehen zuerst verloren?
transport_genes <- c("Slc34a1", "Slc22a6", "Slc5a2", ...)
metabolic_genes <- c("Pck1", "Fbp1", "G6pc", ...)
reabsorption_genes <- c("Lrp2", "Cubn", "Slc7a9", ...)

score_transport <- colMeans(bulk_expr[transport_genes, ])
score_metabolic <- colMeans(bulk_expr[metabolic_genes, ])

```





# Conclusion

#### For 1. PT-specifically expressed genes: 
massive downregulation (36 genes vs only 3 upregulated) indicates substantial loss of proximal tubule identity.
Lost PT transport functions:
•	Phosphate reabsorption: Slc34a1, Slc34a3 (sodium-phosphate cotransporters - hallmark PT markers)
•	Glucose reabsorption: Slc5a2 (SGLT2 - the primary glucose transporter)
•	Amino acid transport: Slc6a18, Slc7a7, Slc7a13
•	Organic anion/cation transport: Slc22a6, Slc22a12 (OAT1, OAT3 family)
•	Other solutes: Slc17a3, Slc27a2

Lost metabolic specialization:
•	Gluconeogenesis: Fbp1, Pck1 precursor, Aldob
•	Amino acid metabolism: Pah (phenylalanine hydroxylase), Ass1 (urea cycle)
•	Fatty acid oxidation: Amacr, Acsm family (early signs)
•	Xenobiotic metabolism: Multiple Cyp enzymes (2j13, 4b1, 2e1), Ugt2b38, Ces1f
This pattern is consistent with reprogramming-induced metabolic rewiring - cells are shifting away from the highly specialized, energy-intensive functions of differentiated PT cells.

After 2 weeks there is an adaptive metabolic reprogramming. Possibly lasting tubular damage (persistently downregulated genes).


#### For 2. Comparison to genes that are all expressed in proximal tubules: 
After 3 days, PT cells are temporarily downregulating their baseline metabolic/transport programs.

This aligns with an early “dedifferentiation-like” response to Yamanaka factor induction — the cells are starting to loosen their mature identity without yet activating “youthful” programs.

For 2 week timepoint:
- Translation at pre/post-synapse, cytoplasmic translation

- Oxidative phosphorylation (OXPHOS)

- Aldehyde metabolic process

These are indicative of enhanced protein synthesis and energy metabolism, i.e., the cells are functionally revving up.

Interpretation: By 2 weeks, PT cells show a rejuvenation-like signature, with increased translational and mitochondrial programs.


For downregulation: Top GO terms:

These GO terms are largely developmental/morphogenetic programs, not core PT metabolic programs.

Interpretation: At 2 weeks, PT cells are reducing developmental signaling, possibly suppressing “non-PT-specific” programs or noise, while rejuvenation/metabolic functions take over.


## However
Even though the PT-specific 2-week upregulated GO terms are OXPHOS/translation, this does not necessarily mean rejuvenation in the broader context:

These are PT-specific genes only, not all genes.

Global 2-week results showed enhanced genes corresponding to pro-aging processes (maybe stress response, inflammation, ECM remodeling).

So what looks like “rejuvenation” in PT-specific metabolism may coexist with overall pro-aging trends in other pathways or cell types.
